Investigating shape standardization using geometric morphometry of pottery from Iron Age northeastern Taiwan

### 2020-04-25 analysis/paper/paper.Rmd
submit to JASR

---
title: "Investigating shape standardization using geometric morphometry of pottery from Iron Age northeastern Taiwan"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  The emergence of ceramic specialization in prehistoric societies is often linked to shifts in the complexity of social structures, because specialized pottery production can reflect craft specialization and the presence of elite control. Previous work on identifying specialization relies on typological or linear metric analysis. Here we demonstrate how to investigate ceramic standardization by analyzing outlines of ceramic vessels. Outline analysis is useful because, unlike more commonly-used landmark analysis methods, it can effectively quantify shape differences for objects that lack distinctive measurement points needed for landmark analysis. We demonstrate this method using pottery from Kiwulan, a large multi-component Iron Age site (AD 1350-1850) in northeastern Taiwan. To measure ceramic specialization, we quantified pottery standardization by analyzing shape variables with reproducible geometric morphometric methods. We computed coefficients of variation (CVs) for shape coefficients obtained by elliptical Fourier analysis to test for shape standardization. We found significant differences in pottery shape and shape standardization that indicate changes in pottery production resulting from contact with mainland Han Chinese groups in northeastern Taiwan. Our case study, which includes an openly available research compendium of R code, represents an innovative application of outline-based methods in geometric morphometry to answer the anthropological questions of craft specialization. 
keywords: |
  Geometric Morphometrics; craft specialization; Iron age; culture contact; ceramic standardization
highlights: |
  1. Iron Age pottery shapes from Kiwulan, Taiwan, were analyzed using elliptical fourier analysis
  2. Statistical test for the equality of coefficients of variation of shape coefficients was applied
  3. Significant differences in pottery shape were identified between culture contact phases
  4. Pottery shapes show high standardization after contact with mainland Han Chinese groups

@@

su_manv_tab <- 
  su_manv %>% 
  mutate_if(is.numeric, round, 4) %>% 
  select(-Df, -`num Df`) %>% 
  rename(Comnparison = rowname,
         `Pillai's trace` = Pillai,
         `Approximate F value` = `approx F`,
         `degrees of freedom` = `den Df`)

>>>>

---
title: "Investigating shape standardization using geometric morphometry of pottery from Iron Age northeastern Taiwan"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  The emergence of ceramic specialization in prehistoric societies is often linked to shifts in the complexity of social structures, because specialized pottery production can reflect craft specialization and the presence of elite control. Previous work on identifying specialization relies on typological or linear metric analysis. Here we demonstrate how to investigate ceramic standardization by analyzing outlines of ceramic vessels. Outline analysis is useful because, unlike more commonly-used landmark analysis methods, it can effectively quantify shape differences for objects that lack distinctive measurement points needed for landmark analysis. We demonstrate this method using pottery from Kiwulan, a large multi-component Iron Age site (AD 1350-1850) in northeastern Taiwan. To measure ceramic specialization, we quantified pottery standardization by analyzing shape variables with reproducible geometric morphometric methods. We computed coefficients of variation (CVs) for shape coefficients obtained by elliptical Fourier analysis to test for shape standardization. We found significant differences in pottery shape and shape standardization that indicate changes in pottery production resulting from contact with mainland Han Chinese groups in northeastern Taiwan. Our case study, which includes an openly available research compendium of R code, represents an innovative application of outline-based methods in geometric morphometry to answer the anthropological questions of craft specialization. 
keywords: |
  Geometric Morphometrics; craft specialization; Iron age; culture contact; ceramic standardization
highlights: |
  1. Iron Age pottery shapes from Kiwulan, Taiwan, were analyzed using elliptical Fourier analysis
  2. Statistical test for the equality of coefficients of variation of shape coefficients was applied
  3. Significant differences in pottery shape were identified between culture contact phases
  4. Pottery shapes show high standardization after contact with mainland Han Chinese groups

@@

su_manv_tab <- 
  su_manv %>% 
  mutate_if(is.numeric, round, 4) %>% 
  select(-Df, -`num Df`) %>% 
  rename(Comparison = rowname,
         `Pillai's trace` = Pillai,
         `Approximate F value` = `approx F`,
         `degrees of freedom` = `den Df`)
